---
title: "ML PostModAssignment2"
author: "Syon Saha"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

## Data Prep
```{r, warning=FALSE, include=FALSE}
library("readxl") # used to read excel files
library("dplyr") # used for data munging 
library("FNN") # used for knn regression (knn.reg function)
library("caret") # used for various predictive models
library("class") # for using confusion matrix function
library("rpart.plot") # used to plot decision tree
library("rpart")  # used for Regression tree
library("glmnet") # used for Lasso and Ridge regression
library('NeuralNetTools') # used to plot Neural Networks
library("PRROC") # top plot ROC curve
library("ROCR") # top plot lift curve
library("tidyverse")
library("xgboost")
library("skimr")

```

Step 1: Loading the dataset
```{r, warning=FALSE}
highnote_df <- read_csv("synthetic_HN_data.csv")

# skim(highnote_df)
```

Step 2: Create x an y data set 
```{r}
highnote_y <- highnote_df %>% pull(adopter)
#skim(highnote_y)

highnote_x <- highnote_df %>% select(-adopter, -net_user)
#skim(highnote_x)
```

Step 3: Normalize function and normalizing x variables 
```{r}
normalize <- function(x) {
  return ((x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)))
}

highnote_x_normalized <- as.data.frame(lapply(highnote_x, normalize)) # Normalize x variables, they are at different scale
#skim(highnote_x_normalized)
```

Step 4: Create training and test data sets 
```{r}
smp_size <- floor(0.75 * nrow(highnote_x_normalized))
set.seed(12345)

train_indices <- sample(seq_len(nrow(highnote_x_normalized)), size = smp_size)

# Create the training set for X by selecting only the rows from train_indices
x_train <- highnote_x_normalized[train_indices, ]
# Create the testing set for X by selecting all rows EXCEPT those in train_indices
x_test <- highnote_x_normalized[-train_indices, ]

# Create the training set for y by selecting the corresponding elements from the vector
y_train <- highnote_y[train_indices]
# Create the testing set for y
y_test <- highnote_y[-train_indices]
```

Step 5: Manually impute the data 
```{r}
# Manually impute the data first to avoid data leakage, also imputing doesn't work when you ALSO SMOTE (?) 
impute_model <- preProcess(x_train, method = "medianImpute")

x_train_imputed <- predict(impute_model, x_train)
# Impute test data using the same model
x_test_imputed <- predict(impute_model, x_test)

# IMPORTANT: For classification, caret works best when the outcome is a 'factor'
y_train <- as.factor(y_train)

#skim(x_train_imputed)
```

Step 6: Create DF to store results 
```{r}
# Create an empty data frame to store results from different models
clf_results <- data.frame(matrix(ncol = 5, nrow = 0))
names(clf_results) <- c("Model", "Accuracy", "Precision", "Recall", "F1")

# Create an empty data frame to store TP, TN, FP and FN values
cost_benefit_df <- data.frame(matrix(ncol = 5, nrow = 0))
names(cost_benefit_df) <- c("Model", "TP", "FN", "FP", "TN")
```


## Logistic Regression

Step 1: (Question 2a) Building a model using only the previous time period’s change variables
```{r}
# Data for Model A (Change Variables Only)

# Identify the names of your "change" columns 
change_variable_names <- c("delta1_friend_cnt", "delta1_avg_friend_age", "delta1_avg_friend_male",
                           "delta1_friend_country_cnt", "delta1_subscriber_friend_cnt",
                           "delta1_songsListened", "delta1_lovedTracks", "delta1_posts",
                           "delta1_playlists", "delta1_shouts", "delta1_good_country")

# Create the new data frames for Model A that contain ONLY these columns
x_train_model_A <- x_train_imputed %>% select(all_of(change_variable_names))
x_test_model_A  <- x_test  %>% select(all_of(change_variable_names))
```

Step 2: Regression using Model A
```{r, warning=FALSE, cache=TRUE}
smote_control <- trainControl(
    method = "cv", 
    number = 10,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    sampling = "smote" # SMOTE
)

y_train_factor <- as.factor(y_train)
levels(y_train_factor) <- make.names(levels(y_train_factor))

y_test_factor <- as.factor(y_test)
levels(y_test_factor) <- make.names(levels(y_test_factor))

model_A_smote <- train(
    x = x_train_model_A,
    y = y_train_factor,
    method = "glm",
    family = "binomial",
    trControl = smote_control,
    metric = "ROC" # Optimize for a better metric than accuracy
)

predictions_A_smote <- predict(model_A_smote, newdata = x_test_model_A)

# Create and print the confusion matrix
confusionMatrix_A_smote <- confusionMatrix(predictions_A_smote, y_test_factor, positive = "X1")

print(confusionMatrix_A_smote)

# Add results to clf_results df
overall_metrics_A <- confusionMatrix_A_smote[["overall"]]
class_metrics_A   <- confusionMatrix_A_smote[["byClass"]]
conf_table_A      <- confusionMatrix_A_smote[["table"]]

# Add a new row to the clf_results dataframe
clf_results[nrow(clf_results) + 1,] <- list(
    Model = "LogReg Model A",
    Accuracy = round(overall_metrics_A[["Accuracy"]], 3),
    Precision = round(class_metrics_A[["Precision"]], 3),
    Recall = round(class_metrics_A[["Recall"]], 3),
    F1 = round(class_metrics_A[["F1"]], 3)
)

# Add a new row to the cost_benefit_df dataframe
cost_benefit_df[nrow(cost_benefit_df) + 1,] <- list(
    Model = "LogReg Model A",
    TP = conf_table_A[2, 2], # True Positives
    FN = conf_table_A[1, 2], # False Negatives
    FP = conf_table_A[2, 1], # False Positives
    TN = conf_table_A[1, 1]  # True Negatives
)
```

Step 3: (Question 2b) Regression using Model B
```{r, warning=FALSE, cache=TRUE}
smote_control <- trainControl(
    method = "cv", 
    number = 10,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    sampling = "smote" 
)

model_B_smote <- train(
    x = x_train_imputed, 
    y = y_train_factor,
    method = "glm",
    family = "binomial",
    trControl = smote_control,
    metric = "ROC" 
)

predictions_B_smote <- predict(model_B_smote, newdata = x_test_imputed)

confusionMatrix_B <- confusionMatrix(predictions_B_smote, y_test_factor, positive = "X1")

print(confusionMatrix_B)

# Create and print the confusion matrix
overall_metrics_B <- confusionMatrix_B[["overall"]]
class_metrics_B   <- confusionMatrix_B[["byClass"]]
conf_table_B      <- confusionMatrix_B[["table"]]

# Add a new row to the clf_results dataframe
clf_results[nrow(clf_results) + 1,] <- list(
    Model = "LogReg Model B",
    Accuracy = round(overall_metrics_B[["Accuracy"]], 3),
    Precision = round(class_metrics_B[["Precision"]], 3),
    Recall = round(class_metrics_B[["Recall"]], 3),
    F1 = round(class_metrics_B[["F1"]], 3)
)

# Add a new row to the cost_benefit_df dataframe
cost_benefit_df[nrow(cost_benefit_df) + 1,] <- list(
    Model = "LogReg Model B",
    TP = conf_table_B[2, 2], # True Positives
    FN = conf_table_B[1, 2], # False Negatives
    FP = conf_table_B[2, 1], # False Positives
    TN = conf_table_B[1, 1]  # True Negatives
)
```
# It is intuitive that by adding MORE variables and current period variables will yield better results; as a result, moving further I will only be running models on all/current period variables as it will save time when I run the following models #

## KNN
Step 1: Running the KNN model 
```{r message=FALSE, cache=TRUE}
cross_validation <- trainControl(## 10-fold CV
                                 method = "repeatedcv",
                                 number = 10,
                                 repeats = 1,
                                 sampling = "smote")

Param_Grid <- expand.grid(k = seq(from = 3, to = 15, by = 2))

# Measure training time
start_time_knn <- Sys.time()

knn_clf_fit <- train(x_train_imputed,
                     y_train_factor, 
                     method = "knn",
                     #preProcess = "medianImpute",
                     tuneGrid = Param_Grid,
                     trControl = cross_validation )

end_time_knn <- Sys.time()

elapsed_time_knn <- end_time_knn - start_time_knn
print(paste("Total training time:", round(elapsed_time_knn, 2)))

knn_clf_fit
```

Step 2: How many neighbors is best 
```{r}
# Plot accuracies for different k values
plot(knn_clf_fit) 

# print the best model
print(knn_clf_fit$finalModel)

#Best model is 3 nearest neighbor
```

Step 3: Confusion Matrix 
```{r}
# Predict on test data
knnPredict <- predict(knn_clf_fit, newdata = x_test_imputed) 

# Print Confusion matrix, Accuracy, Sensitivity etc 
confusionMatrix(knnPredict, y_test_factor, positive = "X1")


# Save results in to clf_results
cm_knn <- confusionMatrix(knnPredict, y_test_factor, positive = "X1")

overall_metrics <- cm_knn[["overall"]]
class_metrics <- cm_knn[["byClass"]]
conf_table <- cm_knn[["table"]]

clf_results[nrow(clf_results) + 1,] <- list(
    Model = "KNN",
    Accuracy = round(overall_metrics[["Accuracy"]], 3),
    Precision = round(class_metrics[["Precision"]], 3),
    Recall = round(class_metrics[["Recall"]], 3),
    F1 = round(class_metrics[["F1"]], 3)
)

cost_benefit_df[nrow(cost_benefit_df) + 1,] <- list(
    Model = "KNN",
    TP = conf_table[2, 2], # True Positives
    FN = conf_table[1, 2], # False Negatives
    FP = conf_table[2, 1], # False Positives
    TN = conf_table[1, 1]  # True Negatives
)
```


## Decision Tree
Step 1: Decision Tree Model
```{r message=FALSE, cache=TRUE}
cross_validation_dt <- trainControl(
                                    method = "repeatedcv",
                                    number = 10,
                                    repeats = 3,
                                    classProbs = TRUE, # Required for SMOTE
                                    summaryFunction = twoClassSummary, 
                                    sampling = "smote" # This is the key line to add SMOTE
                                )

param_grid <- expand.grid(maxdepth = 2:10)

# Measure training time
start_time_dt <- Sys.time()

dtree_fit <- train(x_train_imputed,
                   y_train_factor, 
                   method = "rpart2",
                   # split - criteria to split nodes
                   parms = list(split = "gini"),
                   tuneGrid = param_grid,
                   trControl = cross_validation_dt,
                   metric = "ROC",
                   preProc = c("center", "scale"))

end_time_dt <- Sys.time()

elapsed_time_dt <- end_time_dt - start_time_dt
print(paste("Total training time:", round(elapsed_time_dt, 2)))

# check the accuracy for different models
dtree_fit
```

Step 2
```{r}
# print the final model
dtree_fit$finalModel
```

Step 3: Plot the tree
```{r}
# Plot decision tree
prp(dtree_fit$finalModel, box.palette = "Reds", tweak = 1.2)
```

Step 4: Decision Tree Confusion Matrix 
```{r}
# Predict on test data
dtree_predict <- predict(dtree_fit, newdata = x_test_imputed)

y_test_factor <- as.factor(y_test)
levels(y_test_factor) <- make.names(levels(y_test_factor)) # <-- This is the one-line fix!

conf_matrix_dt <- confusionMatrix(
  data = dtree_predict,
  reference = y_test_factor,
  positive = "X1"
)

print(conf_matrix_dt)

# Save results 
overall_metrics_dt <- conf_matrix_dt[["overall"]]
class_metrics_dt   <- conf_matrix_dt[["byClass"]]
conf_table_dt      <- conf_matrix_dt[["table"]]

# Add a new row to the clf_results dataframe
clf_results[nrow(clf_results) + 1,] <- list(
    Model = "Decision Tree",
    Accuracy = round(overall_metrics_dt[["Accuracy"]], 3),
    Precision = round(class_metrics_dt[["Precision"]], 3),
    Recall = round(class_metrics_dt[["Recall"]], 3),
    F1 = round(class_metrics_dt[["F1"]], 3)
)

# Add a new row to the cost_benefit_df dataframe
cost_benefit_df[nrow(cost_benefit_df) + 1,] <- list(
    Model = "Decision Tree",
    TP = conf_table_dt[2, 2], # True Positives
    FN = conf_table_dt[1, 2], # False Negatives
    FP = conf_table_dt[2, 1], # False Positives
    TN = conf_table_dt[1, 1]  # True Negatives
)

```


## XGBoost
Step 1
```{r message=FALSE,  warning=FALSE, cache=TRUE}
# This tells caret how to handle the training process, including the sampling
smote_control <- trainControl(
    method = "cv",
    number = 10,
    classProbs = TRUE,                 # Required for SMOTE
    summaryFunction = twoClassSummary, # Use metrics better suited for imbalance
    sampling = "smote"                 # This enables SMOTE
)

# Measure training time
start_time_xg <- Sys.time()

XG_clf_fit_hn <- train(
    x_train_imputed,
    y_train_factor, 
    method = "xgbTree",
    trControl = smote_control, # <-- Use the control object with SMOTE
    metric = "ROC", # <-- Optimize for a better metric
    preProc = c("center", "scale")
)

end_time_xg <- Sys.time()

# Print elapsed time
elapsed_time_xg <- end_time_xg - start_time_xg
print(paste("Total training time:", round(elapsed_time_xg, 2)))
```

Step 2
```{r }
# print the final model
XG_clf_fit_hn$finalModel
```

Step 3 
```{r }
# Predict on test data
XG_clf_predict_hn <- predict(XG_clf_fit_hn,x_test)
```

Step 4: XGBoost Confusion matrix
```{r }
# Print Confusion matrix, Accuracy, Sensitivity etc 
confusionMatrix(XG_clf_predict_hn, y_test_factor, positive = "X1" )


cm_xg <- confusionMatrix(XG_clf_predict_hn, y_test_factor, positive = "X1" )

overall_metrics_xg <- cm_xg[["overall"]]
class_metrics_xg   <- cm_xg[["byClass"]]

clf_results[nrow(clf_results) + 1,] <- list(
    Model = "XGBoost",
    Accuracy = round(overall_metrics_xg[["Accuracy"]], 3),
    Precision = round(class_metrics_xg[["Precision"]], 3),
    Recall = round(class_metrics_xg[["Recall"]], 3),
    F1 = round(class_metrics_xg[["F1"]], 3)
)

# Extract the confusion matrix table
conf_table_xg <- cm_xg[["table"]]

cost_benefit_df[nrow(cost_benefit_df) + 1,] <- list(
    Model = "XGBoost",
    TP = conf_table_xg[2, 2], # True Positives
    FN = conf_table_xg[1, 2], # False Negatives
    FP = conf_table_xg[2, 1], # False Positives
    TN = conf_table_xg[1, 1]  # True Negatives
)

```


## Neural Network
Step 1
```{r  cache=TRUE, Emessage=FALSE}
smote_control <- trainControl(
    method = "cv",
    number = 10,
    classProbs = TRUE,                 # Required for SMOTE
    summaryFunction = twoClassSummary, # Use metrics better suited for imbalance
    sampling = "smote"                 # This enables SMOTE
)

# Try different combinations of parameters like decay (prevents the weights from growing too large,) and size of Hidden layers
my.grid <- expand.grid(.decay = c(0.1, 0.01), .size = c(5, 7, 9))

# Measure training time
start_time_nn <- Sys.time()

# stepmax is maximum steps for the training of the neural network
# threshold is set to 0.01, meaning that if the change in error during an iteration is 
# less than 1%, then no further optimization will be carried out by the model
nn_clf_fit <- train(
    x_train_imputed,
    y_train_factor,
    method = "nnet",
    trControl = smote_control, # <-- Use the control object with SMOTE
    tuneGrid = my.grid,
    metric = "ROC", # <-- Optimize for a better metric
    trace = FALSE,
    linout = 0,
    stepmax = 100,
    threshold = 0.01
  )

print(nn_clf_fit)

end_time_nn <- Sys.time()

# Print elapsed time
elapsed_time_nn <- end_time_nn - start_time_nn
print(paste("Total training time:", round(elapsed_time_nn, 2)))

# Plot Neural Network 
plotnet(nn_clf_fit$finalModel, 
        y_names = "Premium Adopter",
        max_sp = 1,
        circle_cex = 2.5,
        cex_val = 0.6)

```

Step 2
```{r }
# Predict on test data
nn_clf_predict <- predict(nn_clf_fit,x_test_imputed)
```

Step 3: Confusion matrix
```{r }
# Print Confusion matrix, Accuracy, Sensitivity etc 
confusionMatrix(nn_clf_predict, y_test_factor, positive = "X1")


cm_nn <- confusionMatrix(nn_clf_predict, y_test_factor, positive = "X1")

# save results to df
overall_metrics_nn <- cm_nn[["overall"]]
class_metrics_nn   <- cm_nn[["byClass"]]

clf_results[nrow(clf_results) + 1,] <- list(
    Model = "Neural Net",
    Accuracy = round(overall_metrics_nn[["Accuracy"]], 3),
    Precision = round(class_metrics_nn[["Precision"]], 3),
    Recall = round(class_metrics_nn[["Recall"]], 3),
    F1 = round(class_metrics_nn[["F1"]], 3)
)

# Extract the confusion matrix table
conf_table_nn <- cm_nn[["table"]]

cost_benefit_df[nrow(cost_benefit_df) + 1,] <- list(
    Model = "Neural Net",
    TP = conf_table_nn[2, 2], # True Positives
    FN = conf_table_nn[1, 2], # False Negatives
    FP = conf_table_nn[2, 1], # False Positives
    TN = conf_table_nn[1, 1]  # True Negatives
)
```


## Random Forest  
Step 1: create the model
```{r cache=TRUE}
smote_control_rf <- trainControl(
    method = "cv",
    number = 10,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    sampling = "smote"
)

rf_param_grid <- expand.grid(mtry = c(2, 5, 10))

# Measure training time
start_time_rf <- Sys.time()

rf_fit_smote <- train(
    x = x_train_imputed,
    y = y_train_factor,
    method = "rf",
    metric = "ROC",
    trControl = smote_control_rf,
    tuneGrid = rf_param_grid,
)

end_time_rf <- Sys.time()

# Print elapsed time
elapsed_time_rf <- end_time_rf - start_time_rf
print(paste("Total training time:", round(elapsed_time_rf, 2)))

print(rf_fit_smote)
```

Step 2: Predict on the test data
```{r}
# Predict on the test data
rf_predict <- predict(rf_fit_smote, newdata = x_test_imputed)
```

Step 3: Confusion Matrix 
```{r}
# Create the confusion matrix
cm_rf <- confusionMatrix(rf_predict, y_test_factor, positive = "X1")
print(cm_rf)

# Extract metrics and add to your results dataframes
overall_metrics_rf <- cm_rf[["overall"]]
class_metrics_rf   <- cm_rf[["byClass"]]
conf_table_rf      <- cm_rf[["table"]]

# Add to clf_results
clf_results[nrow(clf_results) + 1,] <- list(
    Model = "Random Forest",
    Accuracy = round(overall_metrics_rf[["Accuracy"]], 3),
    Precision = round(class_metrics_rf[["Precision"]], 3),
    Recall = round(class_metrics_rf[["Recall"]], 3),
    F1 = round(class_metrics_rf[["F1"]], 3)
)

# Add to cost_benefit_df
cost_benefit_df[nrow(cost_benefit_df) + 1,] <- list(
    Model = "Random Forest",
    TP = conf_table_rf[2, 2],
    FN = conf_table_rf[1, 2],
    FP = conf_table_rf[2, 1],
    TN = conf_table_rf[1, 1]
)
```


## Results 
```{r}
print(cost_benefit_df)
print(clf_results)
```
## Based on the following models and their results, there are a few models that stand out to me. XGBoost has the highest accuracy of 93% but also has the lowest recall out of all the models, 6.5%. XGBoost is highly conservative and sorts most users as TN which causes the model to have the lowest TP count of 120. In other words, it fails to identify more users who would have converted. The purpose of this assignment/dataset is to identify the top 1000 users who are most likely to convert. XGBoost, while having the best accuracy, is not the best model at targeting users to convert.
## I have chosen Random Forest as the best model. This is true because we have a business constraint on how many users we want to target, 1000. When generating a fixed list like this, the model's Precision is the most important metric because it tells us how "pure" the list of predicted converters is. The Random Forest model has a Precision of 28.4%, which is significantly better than any other model, including the next best, Logistic Regression (17.8%). This means the Random Forest will predict users who are more likely to convert. While other models like the Neural Net have higher recall, the goal is to find the best top 1000, and the Random Forest model does that the best due to its high precision. 

## Top 1000 users to target with the free-trial 
```{r}
# Define final model 
final_model <- rf_fit_smote


#Impute the highnote df 
highnote_df_imputed <- predict(impute_model, highnote_df)

# Predict the probability for the positive class ("X1" or "adopter")
# The 'type = "prob"' argument is essential here.
predictions_full <- predict(final_model, newdata = highnote_df_imputed, type = "prob")

# Add the prediction probabilities as a new column to your original data
results_df <- highnote_df
# We select the second column of the prediction output, which is the probability of class "X1"
results_df$conversion_probability <- predictions_full$X1

# Filter for current free users, sort by probability, and select the top 1000
top_1000_target_list <- results_df %>%
  filter(adopter == 0) %>% # Only target current FREE users
  arrange(desc(conversion_probability)) %>% # Sort by probability, highest first
  head(1000) %>% # Select the top 1000 rows
  select(net_user, conversion_probability) # Keep only the user ID and their probability

print(top_1000_target_list)
```


